Deciding Bitvector Termination with Program Synthesis

### 2014-07-03 papers/termination/paper.tex
fixed inconsistencies in sec 5

We will begin our discussion by showing how to encode the (non-)termination of a single loop
in the synthesis fragment.  For the time being, a loop is a pair ${L= \langle G,T \rangle}$,
where the states $x$ satisfying the loop's guard are given by the predicate $G(x)$.
The body of the loop is encoded as the transition relation $T(x, x')$, meaning that
state $x'$ is reachable from state $x$ via a single iteration of the loop body.
For example, the loop in Figure~\ref{fig:motivation.a} is encoded as:
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}
We will abbreviate this with the notation:
\begin{align*}
G(x) & \triangleq x > 0 \\
T(x, x') & \triangleq x' = (x - 1) \, \& \, x
\end{align*}

@@

\begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
\label{def:nonterm-formula}
 \begin{align*}
  \exists N, x_0 . \forall x . \exists x' . & I(x_0) \wedge N(x_0) ~ \wedge \\
                                                        & N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \rightarrow T(x, x') \wedge N(x') 
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
\label{def:deterministic-nonterm-formula}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & I(x_0) \wedge N(x_0) ~ \wedge \\
                                                        & N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T(x, x') \rightarrow N(x') 
 \end{align*}
\end{definition}

@@

Continuing the example of the program in Figure~\ref{fig:motivation.a}, we can see that
the corresponding synthesis formula {\bf [UT]} is satisfiable, as witnessed by the function $R(x) = x$.
So $R(x) = x$ constitutes a proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
terminating.

\noindent{\bf Non-termination.}
Dually to termination, we might want to consider the non-termination of a program.  If a program terminates,
we can prove this by finding a ranking function and supporting invariant witnessing the satisfiability
of formula {\bf[CT]}.  What then would a proof of non-termination look like?

Gupta et al. \cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination of a transition relation $T$ by the existence of an \emph{(open) recurrence set},
i.e. a nonempty set of states $N$ such that for each $s \in N$ there
exists a transition to some $s'\in N$. 
For our case, given a loop $L = \langle I,G,T \rangle$, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.

@@

In order to formulate the non-termination problem inside the synthesis fragment, 
we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. in \cite{DBLP:conf/tacas/ChenCFNO14}. 
As opposed to a closed recurrence set, this is a $\exists \forall$ property: for each state in the recurrence set $N$, all of its successors 
must be in $N$. We note that if $T$ is deterministic, every open recurrence set is also a closed recurrence set (since each
state has at most one successor).  Thus, the non-termination problem for deterministic transition systems is equivalent to the
satisfiability of formula {\bf [CNT]}. %Definition~\ref{def:deterministic-nonterm-formula}.
%As our formulation is with respect to a given loop $L=\langle I,G,T \rangle$, we require the  
%recurrence set to include the loop's guard, $N(x) \rightarrow G(x)$. In other words, the loop's guard needs to be satisfied in order for the 
%infinite execution to happen.

@@

 \begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
  \begin{align*}
\label{def:multi-termination-formula}
  \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
 & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
 & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
 & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) \rightarrow Inv_n(x_n) ~ \wedge \\
 & Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n') \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')
 \end{align*}
\end{definition}

\begin{definition}[Nested Loops Termination Formula {\bf [NT]}]
\label{def:nested-term-formula}
 \begin{align*}
  \exists R, S . \forall x, x' . & G_1(x) \wedge P_1(x,x') \rightarrow S(x',x') ~ \wedge \\
                                & G_2(x') \wedge S(x,x') \wedge T(x',x'')\rightarrow S(x,x'') ~ \wedge \\
				& G_1(x) \wedge P_1(x,x') \wedge \neg G_2(x') \wedge S(x',x'') \wedge 
                                P_2(x'',x''') \wedge 
                                  R(x) > 0 \wedge R(x) > R(x''') 
 \end{align*} 
\end{definition}

\begin{definition}[Sequential Loops Non-Termination Formula {\bf [SNT]}]
\label{def:multi-termination-formula}
 \begin{align*}
  \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
 & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
 & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
 & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge N(x_n)  ~ \wedge \\
                                                        & N(x_n) \rightarrow G_n(x_n) ~ \wedge 
							 N(x_n) \wedge T_n(x_n, x_n') \rightarrow N(x_n') 
 \end{align*}
\end{definition}
 \end{framed}

@@

Sometimes the termination behaviour of a loop depends on its environment.  That is to say,
the loop may not terminate if started in some particular state, but the rest of the program
ensures that such a state cannot be reached on entry to the loop.  The program as a whole
terminates, but if the loop were considered in isolation we would not be able to prove that
it terminates.  We must therefore somehow encode a loop's interaction with the rest of
the program in order to do a sound termination analysis.

@@

However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
$R(\langle x,y\rangle) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
This constitutes a proof that the program as a whole terminates, since the loop always begins
executing in a state that guarantees its termination.

@@

To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
it suffices to summarise the inner loop with the relation:
\begin{align*}
S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
\end{align*}
Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.

>>>>

We will begin our discussion by showing how to encode the (non-)termination of program consisting of a single loop 
in the synthesis fragment.  For the time being, a loop is a pair ${L= \langle G,T \rangle}$,
where the states $x$ satisfying the loop's guard are given by the predicate $G(x)$.
The body of the loop is encoded as the transition relation $T(x, x')$, meaning that
state $x'$ is reachable from state $x$ via a single iteration of the loop body.
For example, the loop in Figure~\ref{fig:motivation.a} is encoded as:
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}
We will abbreviate this with the notation:
\begin{align*}
G(x) & \triangleq x > 0 \\
T(x, x') & \triangleq x' = (x - 1) \, \& \, x
\end{align*}

@@

\begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
\label{def:nonterm-formula}
 \begin{align*}
  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \rightarrow T(x, x') \wedge N(x') 
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
\label{def:deterministic-nonterm-formula}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T(x, x') \rightarrow N(x') 
 \end{align*}
\end{definition}

@@

Continuing the example of the program in Figure~\ref{fig:motivation.a}, we can see that
the corresponding synthesis formula {\bf [UT]} is satisfiable, as witnessed by the function $R(x) = x$.
So $R(x) = x$ constitutes a proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
terminating.\\

\noindent{\bf Non-termination.}
Dually to termination, we might want to consider the non-termination of a program.  If a program terminates,
we can prove this by finding a ranking function %and supporting invariant 
witnessing the satisfiability of formula {\bf[UT]}.  What then would a proof of non-termination look like?

Gupta et al. \cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination of a transition relation $T$ by the existence of an \emph{(open) recurrence set},
i.e. a nonempty set of states $N$ such that for each $s \in N$ there
exists a transition to some $s'\in N$. 
For our case, given a loop $L = \langle G,T \rangle$, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.

@@

In order to formulate the non-termination problem inside the synthesis fragment, 
we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. in \cite{DBLP:conf/tacas/ChenCFNO14}. 
As opposed to an open recurrence set, this is a $\exists \forall$ property: for each state in the recurrence set $N$, all of its successors 
must be in $N$. We note that if $T$ is deterministic, every open recurrence set is also a closed recurrence set (since each
state has at most one successor).  Thus, the non-termination problem for deterministic transition systems is equivalent to the
satisfiability of formula {\bf [CNT]}. %Definition~\ref{def:deterministic-nonterm-formula}.
%As our formulation is with respect to a given loop $L=\langle I,G,T \rangle$, we require the  
%recurrence set to include the loop's guard, $N(x) \rightarrow G(x)$. In other words, the loop's guard needs to be satisfied in order for the 
%infinite execution to happen.

@@

%%  \begin{definition}[Sequential Loops Termination Formula {\bf [ST]}]
%%   \begin{align*}
%% \label{def:multi-termination-formula}
%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge G_n(x_n) \rightarrow Inv_n(x_n) ~ \wedge \\
%%  & Inv_n(x_n) \wedge G_n(x_n) \wedge T(x_n, x_n') \rightarrow Inv_n(x_n') \wedge R(x_n) > 0 \wedge R(x_n) > R(x_n')
%%  \end{align*}
%% \end{definition}

%% \begin{definition}[Nested Loops Termination Formula {\bf [NT]}]
%% \label{def:nested-term-formula}
%%  \begin{align*}
%%   \exists R, S . \forall x, x' . & G_1(x) \wedge P_1(x,x') \rightarrow S(x',x') ~ \wedge \\
%%                                 & G_2(x') \wedge S(x,x') \wedge T(x',x'')\rightarrow S(x,x'') ~ \wedge \\
%% 				& G_1(x) \wedge P_1(x,x') \wedge \neg G_2(x') \wedge S(x',x'') \wedge 
%%                                 P_2(x'',x''') \wedge 
%%                                   R(x) > 0 \wedge R(x) > R(x''') 
%%  \end{align*} 
%% \end{definition}

%% \begin{definition}[Sequential Loops Non-Termination Formula {\bf [SNT]}]
%% \label{def:multi-termination-formula}
%%  \begin{align*}
%%   \exists R, Inv_1,..., Inv_n . \forall x_0,...,x_n, x_1',...,x_n'.  & P_0(x_0,x_1) \rightarrow Inv_1(x_1) ~  \\
%%  & \bigwedge_{i=1..n{-}1} (Inv_i(x_i) \wedge G_i(x_i) \wedge T_i(x_i, x_i') \rightarrow Inv_i(x_i')) ~  \\  
%%  & \bigwedge_{i=1..n{-}2} (Inv_i(x_i) \wedge \lnot G_i(x_i) \wedge P_{i+1}(x_i, x_{i+1}) \rightarrow Inv_{i+1}(x_{i+1})) ~ \wedge \\
%%  & Inv_{n-1}(x_{n-1}) \wedge \lnot G_{n-1}(x_{n-1}) \wedge P_n(x_{n-1},x_n) \wedge N(x_n)  ~ \wedge \\
%%                                                         & N(x_n) \rightarrow G_n(x_n) ~ \wedge 
%% 							 N(x_n) \wedge T_n(x_n, x_n') \rightarrow N(x_n') 
%%  \end{align*}
%% \end{definition}
 \end{framed}

@@

Sometimes the termination behaviour of a loop depends on its environment.  That is to say,
the loop may not terminate if started in some particular state, but the rest of the program
ensures that such a state cannot be reached on entry to the loop.  The program as a whole
terminates, but if the loop were considered in isolation we would not be able to prove that
it terminates.  We must therefore encode a loop's interaction with the rest of
the program in order to do a sound termination analysis.\\

@@

However, formula {\bf [CT]} is satisfiable, as witnessed by the ranking function
$R(\langle x,y\rangle) = x$ and supporting invariant $W(\langle x, y \rangle ) \triangleq y = 1$.
This constitutes a proof that the program as a whole terminates, since the loop always begins
executing in a state that guarantees its termination.\\

@@

To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
it suffices to summarise the inner loop with the relation:
\begin{align*}
S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
\end{align*}
Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.\\
